---
title: "Getting Started with cdfquant01reg"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with cdfquant01reg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(cdfquant01reg)
```

The cdfquant01reg package implements Bayesian regression with the finite-tailed CDF-quantile (FTCDFQ) distributions (Smithson & Shou, 2024) and their extended-support versions in the BRMS package in R.

The package has two families of distributions, the 32 two- and three-parameter FTCDFQ distributions, and the 16 extended-support distributions based on the two-parameter FTCDFQ distributions. This vignette demonstrates the usages of both kinds.

# FTCDFQ Regression

There are two commands for executing a regression model, one identifying the distribution to be used in the regression model, and the other specifying the regression model and the data source. The select_ftcdfq() command displays a menu of 32 FTCDFQ distributions and asks for a number identifying the distribution chosen by the user.

```{r, eval = FALSE}
# Run the interactive menu selection
select_ftcdfq()
```

The function will print out the 32 standard distribution options and prompt you for a choice:

``` text
Choose a BRMS model to load into the session: 

 1: Arcsinh_Arcsinh_outer_W_2   2: Arcsinh_Cauchy_outer_W_2    3: Cauchit_Arcsinh_outer_W_2
...
31: Cauchit_Cauchy_inner_V_3   32: t2_t2_inner_V_3

Selection: 
```

Once you enter a number (e.g., `3`), the custom Stan variables will be registered automatically, and you will see a message like this one:

``` text
Selection: 3
--> --> Successfully loaded: Cauchit_Arcsinh_outer_W_2
The shared and unique Stan functions are now active.
```

The package handles the BRMS custom family and stanvars automatically, so the ftcdfq command for running a model needs only a BRMS formula and data source, although other BRMS control parameters can be added as well (e.g., the number of chains). This example executes an intercepts model using a two-parameter distribution for the dependent variable pphysdecis from a dataset named “yoon” (Example 3 from Smithson and Shou, 2024, see below for a description of the dataset).

```{r, eval = FALSE}
data(yoon)
# Run the model
m1 <- ftcdfq(bf(pphysdecis ~ 1, sigma ~ 1), data=yoon, backend = "cmdstanr")
# Get the model summary
summary(m1)
```

BRMS post-estimation functions and commands are available, such as log_lik or epred. Also available are estimates of the conditional pdf, cdf, and quantile functions for the distribution employed in the regression model. Their commands are dnsity, CDF, and inverse_CDF, respectively. Given the model above, the following commands produce the model distribution’s median and cdf at three points.

```{r, eval = FALSE}
# Get the model coefficients:
b0 <- fixef(m1)["Intercept", "Estimate"]
s0 <- exp(fixef(m1)["sigma_Intercept", "Estimate"])
# Compute the estimated median:
inverse_CDF(0.5,b0,s0)
# Compare it with the sample median:
median(yoon$pphysdecis)
# Inspect the CDF values for 0.25, 0.5, and 0.75 on the pphysdecis scale:
c(CDF(0.25,b0,s0),CDF(0.5,b0,s0),CDF(0.75,b0,s0))
```

These commands use the dnsity function to plot the fitted distribution against a normed histogram of the data.

```{r, eval = FALSE}
library(MASS)
# Draw the fitted distribution and histogram
myseq <- seq(0,1,0.01)
mydensity <- c(rep(0,length(myseq)))
for (i in 1:length(mydensity)) {mydensity[i] <- dnsity(myseq[i],b0,s0)}
truehist(yoon$pphysdecis, nbins=50, xlab = "pphysdecis", ylab = "pdf")
lines(myseq,mydensity, lwd = 2)
```

## Example 3 from Smithson and Shou (2024)

The data are records of 894 patients whose times were recorded for five stages of ED assessment and treatment from a study consisting of recorded information on patients admitted to the emergency department of the University of Alberta Hospital between midnight January 23 and midnight January 29, 1999 (Yoon, Steiner, and Reinhardt 2003). Smithson and Shou used a Cauchit-Cauchy outer-W two-parameter distribution to assess the difference in the proportion of time spent in treatment (versus registration and triage) between patients arriving by ambulance and those who were “walk-ins”. The code here tests their hypothesis that ambulance-arrivals spent more time under treatment by a physician, using the Cauchit-Arcsinh outer-W distribution that already has been employed for this vignette.

```{r, eval = FALSE}
# Run the model
m2 <- ftcdfq(bf(pphysdecis ~ Ambulance, sigma ~ Ambulance), data=yoon, backend = "cmdstanr")
# Get the model summary
summary(m2)
# Get the model log-likelihood:
LL1 <- log_lik(m1)
sum(colMeans(LL1))
# 
# Get the model estimated medians for the Walk-ins
b0 <- fixef(m2)["Intercept", "Estimate"]
s0 <- exp(fixef(m2)["sigma_Intercept", "Estimate"])
inverse_CDF(0.5,b0,s0)
# Compare it with the sample median:
median(yoon$pphysdecis[yoon$Ambulance==0])
#
# And for the Ambulance Arrivals:
b1 <- fixef(m2)["Ambulance", "Estimate"]
s1 <- exp(fixef(m2)["sigma_Ambulance", "Estimate"])
inverse_CDF(0.5,b0+b1,s0+s1)
# Compare it with the sample median:
median(yoon$pphysdecis[yoon$Ambulance==1])
```

The next code chunk draws the fitted distributions and histograms for walk-ins and ambulance-arrivals.

```{r, eval = FALSE}
# Draw the fitted distribution and histogram for the Walk-ins
myseq <- seq(0,1,0.01)
mydensity <- c(rep(0,length(myseq)))
for (i in 1:length(mydensity)) {mydensity[i] <- dnsity(myseq[i],b0,s0)}
truehist(yoon$pphysdecis[yoon$Ambulance == 0], nbins=50, xlab = "pphysdecis", ylab = "pdf")
lines(myseq,mydensity, lwd = 2)
#
# Draw the fitted distribution and histogram for the Ambulance Arrivals
for (i in 1:length(mydensity)) {mydensity[i] <- dnsity(myseq[i],b0+b1,s0+s1)}
truehist(yoon$pphysdecis[yoon$Ambulance == 1], nbins=50, xlab = "pphysdecis", ylab = "pdf")
lines(myseq,mydensity, lwd = 2)
```

# Extended-Support Regression

This section describes regression modeling for the extended-support FTCDFQ distributions. As before, there are two commands for executing a regression model, one identifying the distribution to be used in a regression model, and the other specifying the regression model and the data source. The select_exftcdfq() command displays a menu of the 16 extended-support FTCDFQ distributions and asks for a number identifying the distribution chosen by the user.

```{r, eval = FALSE}
# Run the interactive menu selection
select_exftcdfq()
```

The function will print out the 16 extended-support distribution options and prompt you for a choice:

``` text
Choose a BRMS model to load into the session: 

 1: Ex_Arcsinh_Arcsinh_outer_W   2: Ex_Arcsinh_Cauchy_outer_W    
 ...
 15: Ex_Cauchit_Cauchy_inner_V   16: Ex_t2_t2_inner_V

Selection: 
```

As before, when you enter a number (e.g., `1`), the custom Stan variables will be registered automatically, and you will see a message like this one:

``` text
Selection: 1
--> Successfully loaded: Ex_Arcsinh_Arcsinh_outer_W
The shared and unique Stan functions are now active.
```

This example first executes an intercepts model using the selected distribution for the dependent variable ptotpalest1 from a dataset named “blame” (see below for a description of the dataset). The u_Intercept in the model summary is the logged exceedance parameter, which is the amount by which the unit interval has to be extended above 1 and below 0 to handle the zeros and ones in the data.

```{r, eval = FALSE}
data(blame)
# Run the model
m3 <- exftcdfq(bf(ptotpalest1 ~ 1, sigma ~ 1, u ~ 1), data= blame)
# Get the model summary
summary(m3)
# Get the model log-likelihood:
LL3 <- log_lik(m3)
sum(colMeans(LL3))
#
# model distribution’s median:
b0 <- fixef(m1)["Intercept", "Estimate"]
s0 <- exp(fixef(m1)["sigma_Intercept", "Estimate"])
u0 <- exp(fixef(m1)["u_Intercept", "Estimate"])
inverse_CDF(0.5,b0,s0,u0)
# Compare the estimate with the empirical median:
median(blame$ptotpalest1)
# 
```

## Predicting Palestine Blame

The data for this example are from a replication and extension of Halevy et al. (2022), the replication conducted by Smithson. An Israeli adult sample was recruited using the Qualtrics panel platform during April 2021 (N = 462; mean age = 39.9, age SD = 15.5; 245 females, 214 males, 3 other). The participants were asked to assign percentages of blame to Israelis and Palestinians for the unresolved Israeli-Palestinian conflict:

> “At present, a permanent solution to the Israeli-Palestinian conflict has yet to be found. Who do you blame for the current absence of a permanent solution? In the table below you will find two groups. Please indicate how much you think each of them is to blame for the absence of a permanent solution to the conflict. Please note – the total sum of percentages must add up to 100%.”

Two experimental conditions varied whether the Palestinians were listed as one population or “unpacked” into three separate movements: Palestinians listed as “The Palestinians” vs listed as “The Palestinian Hamas movement, The Palestinian Fatah movement, The Palestinian Islamic Jihad movement”. The hypothesis to be tested here was that unpacking into three groups would increase the percentage of blame that the Israeli participants assigned to Palestinians. The regression model below shows support for this hypothesis. The palestpk variable is coded 0 (unpacked) and 1 (packed). The positive effect of palestpk on the skew parameter implies that the unpacked condition yielded a higher median than the packed condition.

```{r, eval = FALSE}
# Run the model
m4 <- exftcdfq(bf(ptotpalest1 ~ palestpk, sigma ~ 1, u ~ 1), data= blame)
# Get the model summary
summary(m4)
# 
# Get the model log-likelihood:
LL4 <- log_lik(m4)
sum(colMeans(LL4))
#
# Get the model estimated medians for the “unpacked” condition
b0 <- fixef(m2)["Intercept", "Estimate"]
s0 <- exp(fixef(m2)["sigma_Intercept", "Estimate"])
u0 <- exp(fixef(m2)["u_Intercept", "Estimate"])
inverse_CDF(0.5,b0,s0,u0)
# Compare the estimate with the empirical median:
median(blame$ptotpalest1[blame$palestpk==0])
#
# And for the “packed” condition:
b1 <- fixef(m2)["palestpk", "Estimate"]
inverse_CDF(0.5,b0+b1,s0,u0)
# Compare the estimate with the empirical median:
median(blame$ptotpalest1[blame$palestpk==1])
# 
# Draw the fitted distribution and histogram for the “packed”
myseq <- seq(-u0,1+u0,0.01)
mydensity <- c(rep(0,length(myseq)))
for (i in 1:length(mydensity)) {mydensity[i] <- dnsity(myseq[i],b0,s0,u0)}
truehist(blame$ptotpalest1[blame$palestpk == 0], nbins=20, xlab = "ptotpalest1", ylab = "pdf", xlim = c(-u0,1+u0))
lines(myseq,mydensity, lwd = 2)
#
# Draw the fitted distribution and histogram for the “unpacked”
for (i in 1:length(mydensity)) {mydensity[i] <- dnsity(myseq[i],b0+b1,s0,u0)}
truehist(blame$ptotpalest1[blame$palestpk == 1], nbins=20, xlab = "ptotpalest1", ylab = "pdf", xlim = c(-u0,1+u0))
lines(myseq,mydensity, lwd = 2)
```

# References

Halevy, N., Maoz, I., Vani, P. and Reit, E.S., 2022. Where the blame lies: Unpacking groups into their constituent subgroups shifts judgments of blame in intergroup conflict. Psychological Science, 33(1), pp.76-89.

Smithson, M. and Shou, Y., 2024. Flexible CDF-quantile distributions on the closed unit interval, with software and applications. Communications in Statistics-Theory and Methods, 53(11), pp.3876-3898.

Yoon, P., I. Steiner, and G. Reinhardt. 2003. Analysis of factors influencing length of stay in the emergency department. CJEM 5 (3):155–61.
