---
title: "Getting Started with cdfquant01reg"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with cdfquant01reg}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(cdfquant01reg)
```

The cdfquant01reg package implements Bayesian regression with the finite-tailed CDF-quantile (FTCDFQ) distributions (Smithson & Shou, 2024) and their extended-support versions in the BRMS package in R. The package has three subfamilies of distributions: the 16 two-parameter distributions, the 16 three-parameter distributions, and 16 extended-support distributions based on the two-parameter FTCDFQ distributions.

# FTCDFQ Regression

There are two kinds of commands for executing a regression model in this package, one kind identifying the distribution to be used in the regression model, and the other kind specifying the regression model and the data source. There is a distribution selection command for each of the three subfamilies: select_ftcdfq2, select_ftcdfq3, and select_exftcdfq. Each of them invites the user to select from a menu of 16 distributions.

```{r, eval = FALSE}
# Run the interactive menu selection
select_ftcdfq2()
```

For instance, the select_ftcdfq2() command displays a menu of the 16 two-parameter distributions and asks for a number identifying the distribution you wish to choose.

``` text
Choose a distribution to load into the session: 

 1: Arcsinh_Arcsinh_outer_W_2   2: Arcsinh_Cauchy_outer_W_2 
 3: Cauchit_Arcsinh_outer_W_2   4: Cauchit_Cauchy_outer_W_2 
 5: t2_t2_outer_W_2             6: Cauchit_Arcsinh_outer_V_2
 7: Cauchit_Cauchy_outer_V_2    8: t2_t2_outer_V_2          
 9: Arcsinh_Arcsinh_inner_W_2  10: Arcsinh_Cauchy_inner_W_2 
11: Cauchit_Cauchy_inner_W_2   12: t2_t2_inner_W_2          
13: Arcsinh_Cauchy_inner_V_2   14: Cauchit_Arcsinh_inner_V_2
15: Cauchit_Cauchy_inner_V_2   16: t2_t2_inner_V_2

Selection: 
```

Once you enter a number (e.g., `3`), the custom Stan variables will be registered automatically, and you will see a message like this one:

``` text
Selection: 3
--> Successfully loaded: Cauchit_Arcsinh_outer_W_2
The shared and unique Stan functions are now active.
```

The package handles the BRMS custom family and stanvars automatically, so the ftcdfq command for running a model needs only a BRMS formula and data source, although other BRMS control parameters can be added as well (e.g., the number of chains). The same is true for a model using maximum likelihood estimation and frequentist inference.

This example begins with an intercepts model using a two-parameter distribution for the dependent variable pphysdecis from a dataset named “yoon” (Example 3 from Smithson and Shou, 2024). The data are records of 894 patients whose times were recorded for five stages of ED assessment and treatment from a study consisting of recorded information on patients admitted to the emergency department of the University of Alberta Hospital between midnight January 23 and midnight January 29, 1999 (Yoon, Steiner, and Reinhardt 2003). The variable pphysdecis is the proportion of the patient's time in the emergency department consisting of consultation with or treatment by a physician.

We begin by fitting the two-parameter Cauchit_Arcsinh_outer_W_2 to the pphysdecis distribution, using BRMS.

```{r, eval = FALSE}
data(yoon)
# Run the model
m1 <- ftcdfq(bf(pphysdecis ~ 1, sigma ~ 1), data=yoon, backend = "cmdstanr", silent = 1)
# Get the model summary
summary(m1)
```

BRMS post-estimation functions and commands are available, such as log_lik or epred. Also available are estimates of the conditional pdf, cdf, and quantile functions for the distribution employed in the regression model. Their commands are dnsity, CDF, and inverse_CDF, respectively. Given the model above, the following commands produce the model distribution’s median and cdf at three points.

```{r, eval = FALSE}
# Get the model coefficients:
b0 <- fixef(m1)["Intercept", "Estimate"]
s0 <- exp(fixef(m1)["sigma_Intercept", "Estimate"])
# Compute the estimated median:
inverse_CDF(0.5,b0,s0)
# Alternatively, 
pred1 <- posterior_predict(m1)
median(pred1)
# Compare them with the sample median:
median(yoon$pphysdecis)
# Inspect the CDF values for 0.25, 0.5, and 0.75 on the pphysdecis scale:
c(CDF(0.25,b0,s0),CDF(0.5,b0,s0),CDF(0.75,b0,s0))
```

The next group of commands uses the dnsity function to plot the fitted distribution against a normed histogram of the data.

```{r, eval = FALSE}
library(MASS)
# Draw the fitted distribution and histogram
myseq <- seq(0,1,0.01)
mydensity <- c(rep(0,length(myseq)))
for (i in 1:length(mydensity)) {mydensity[i] <- dnsity(myseq[i],b0,s0)}
truehist(yoon$pphysdecis, nbins=50, xlab = "pphysdecis", ylab = "pdf")
lines(myseq,mydensity, lwd = 2)
```

Now we run the same model using frequentist methods. The commands ftcdfqMLE2 and ftcdfqMLE3 run maximum likelihood regression models for the two-parameter and three-parameter distributions, respectively. These return objects that use S3 methods for extracting coefficients and summary statistics.

Given that the BRMS model used the default priors, the maximum-likelihood model produces quite similar results.

```{r, eval = FALSE}
# Run the model
m1b <- ftcdfqMLE2(pphysdecis ~ 1|1, data=yoon)
# Get the model summary
summary(m1b)
# Compute the estimated median:
mean(predict(m1b))
# Compare it with the BRMS estimated median:
median(pred1)
#
```

## Example 3 from Smithson and Shou (2024)

Smithson and Shou used a Cauchit-Cauchy outer-W two-parameter distribution to assess the difference in the proportion of time spent in treatment (versus registration and triage) between patients arriving by ambulance and those who were “walk-ins”. The code here tests their hypothesis that ambulance-arrivals spent more time under treatment by a physician, using the Cauchit-Arcsinh outer-W two-parameter distribution we have employed thus far.

```{r, eval = FALSE}
# Run the model
m2 <- ftcdfq(bf(pphysdecis ~ Ambulance, sigma ~ Ambulance), data=yoon, backend = "cmdstanr", silent = 1)
# Get the model summary
summary(m2)
# Get the model log-likelihood:
LL2 <- log_lik(m2)
sum(colMeans(LL2))
# 
# Get the model estimated medians for the Walk-ins
b0 <- fixef(m2)["Intercept", "Estimate"]
s0 <- fixef(m2)["sigma_Intercept", "Estimate"]
inverse_CDF(0.5,b0,exp(s0))
# Alternatively, 
ambo <- data.frame(Ambulance = c(0,1))
pred2 <- posterior_predict(m2, newdata = ambo)
median(pred2[,1])
# Compare it with the sample median:
median(yoon$pphysdecis[yoon$Ambulance==0])
#
# And for the Ambulance Arrivals:
b1 <- fixef(m2)["Ambulance", "Estimate"]
s1 <- fixef(m2)["sigma_Ambulance", "Estimate"]
inverse_CDF(0.5,b0+b1,exp(s0+s1))
# Alternatively, 
median(pred2[,2])
# Compare it with the sample median:
median(yoon$pphysdecis[yoon$Ambulance==1])
```

The next code chunk draws the fitted distributions and histograms for walk-ins and ambulance-arrivals.

```{r, eval = FALSE}
# Draw the fitted distribution and histogram for the Walk-ins
myseq <- seq(0,1,0.01)
mydensity <- c(rep(0,length(myseq)))
for (i in 1:length(mydensity)) {mydensity[i] <- dnsity(myseq[i],b0,s0)}
truehist(yoon$pphysdecis[yoon$Ambulance == 0], nbins=50, xlab = "pphysdecis", ylab = "pdf")
lines(myseq,mydensity, lwd = 2)
#
# Draw the fitted distribution and histogram for the Ambulance Arrivals
for (i in 1:length(mydensity)) {mydensity[i] <- dnsity(myseq[i],b0+b1,s0+s1)}
truehist(yoon$pphysdecis[yoon$Ambulance == 1], nbins=50, xlab = "pphysdecis", ylab = "pdf")
lines(myseq,mydensity, lwd = 2)
```

As before, we run a maximum likelihood model for comparison with the BRMS regression. Again, it produces quite similar results to the BRMS model.

```{r, eval = FALSE}
# Run the model
m2b <- ftcdfqMLE2(pphysdecis ~ Ambulance|Ambulance, data=yoon)
# Get the model summary
summary(m2b)
# Get the model estimated medians for the Ambulancw arrivals and Walk-ins
predict(m2b, newdata = c(0, 1))
# Compare it with the BRMS estimated medians:
c(median(pred2[,1]),median(pred2[,2]))
#
```

# Extended-Support Regression

This section describes regression modeling for the extended-support FTCDFQ distributions. As before, there are two commands for executing a regression model, one identifying the distribution to be used in a regression model, and the other specifying the regression model and the data source. The select_exftcdfq() command displays a menu of the 16 extended-support FTCDFQ distributions and asks for a number identifying the distribution chosen by the user.

```{r, eval = FALSE}
# Run the interactive menu selection
select_exftcdfq()
```

The function will print out the 16 extended-support distribution options and prompt you for a choice:

``` text
Choose a BRMS model to load into the session: 

 1: Ex_Arcsinh_Arcsinh_outer_W   2: Ex_Arcsinh_Cauchy_outer_W    
 ...
 15: Ex_Cauchit_Cauchy_inner_V   16: Ex_t2_t2_inner_V

Selection: 
```

As before, when you enter a number (e.g., `1`), the custom Stan variables will be registered automatically, and you will see a message like this one:

``` text
Selection: 1
--> Successfully loaded: Ex_Arcsinh_Arcsinh_outer_W
The shared and unique Stan functions are now active.
```

This example first executes an intercepts model using the selected distribution for the dependent variable ptotpalest1 from a dataset named “blame” (see below for a description of the dataset). The u_Intercept in the model summary is the logged exceedance parameter, which is the amount by which the unit interval has to be extended above 1 and below 0 to handle the zeros and ones in the data.

```{r, eval = FALSE}
data(blame)
# Run the model
m3 <- exftcdfq(bf(ptotpalest1 ~ 1, sigma ~ 1, u ~ 1), data= blame, backend = "cmdstanr", silent = 1)
# Get the model summary
summary(m3)
# Get the model log-likelihood:
LL3 <- log_lik(m3)
sum(colMeans(LL3))
#
# model distribution’s median:
b0 <- fixef(m3)["Intercept", "Estimate"]
s0 <- exp(fixef(m3)["sigma_Intercept", "Estimate"])
u0 <- exp(fixef(m3)["u_Intercept", "Estimate"])
inverse_CDF(0.5,b0,s0,u0)
# Alternatively,
pred3 <- posterior_predict(m3)
median(pred3)
# Compare the estimate with the empirical median:
median(blame$ptotpalest1)
# 
```

## Predicting Palestine Blame

The data for this example are from a replication and extension of Halevy et al. (2022), the replication conducted by Smithson. An Israeli adult sample was recruited using the Qualtrics panel platform during April 2021 (N = 462; mean age = 39.9, age SD = 15.5; 245 females, 214 males, 3 other). The participants were asked to assign percentages of blame to Israelis and Palestinians for the unresolved Israeli-Palestinian conflict:

> “At present, a permanent solution to the Israeli-Palestinian conflict has yet to be found. Who do you blame for the current absence of a permanent solution? In the table below you will find two groups. Please indicate how much you think each of them is to blame for the absence of a permanent solution to the conflict. Please note – the total sum of percentages must add up to 100%.”

Two experimental conditions varied whether the Palestinians were listed as one population or “unpacked” into three separate movements: Palestinians listed as “The Palestinians” vs listed as “The Palestinian Hamas movement, The Palestinian Fatah movement, The Palestinian Islamic Jihad movement”. The hypothesis to be tested here was that unpacking into three groups would increase the percentage of blame that the Israeli participants assigned to Palestinians. The regression model below shows support for this hypothesis. The palestpk variable is coded 0 (unpacked) and 1 (packed). The positive effect of palestpk on the skew parameter implies that the unpacked condition yielded a higher median than the packed condition.

```{r, eval = FALSE}
# Run the model
m4 <- exftcdfq(bf(ptotpalest1 ~ palestpk, sigma ~ 1, u ~ 1), data= blame, backend = "cmdstanr", silent = 1)
# Get the model summary
summary(m4)
# 
# Get the model log-likelihood:
LL4 <- log_lik(m4)
sum(colMeans(LL4))
#
# Get the model estimated medians for the “unpacked” condition
b0 <- fixef(m4)["Intercept", "Estimate"]
s0 <- exp(fixef(m4)["sigma_Intercept", "Estimate"])
u0 <- exp(fixef(m4)["u_Intercept", "Estimate"])
inverse_CDF(0.5,b0,s0,u0)
# Alternatively, 
pkunpk <- data.frame(palestpk = c(0,1))
pred4 <- posterior_predict(m4, newdata = pkunpk)
median(pred4[,1])
# Compare the estimate with the empirical median:
median(blame$ptotpalest1[blame$palestpk==0])
#
# And for the “packed” condition:
b1 <- fixef(m4)["palestpk", "Estimate"]
inverse_CDF(0.5,b0+b1,s0,u0)
# Alternatively,
median(pred4[,2])
# Compare the estimate with the empirical median:
median(blame$ptotpalest1[blame$palestpk==1])
# 
# Draw the fitted distribution and histogram for the “packed”
myseq <- seq(-u0,1+u0,0.01)
mydensity <- c(rep(0,length(myseq)))
for (i in 1:length(mydensity)) {mydensity[i] <- dnsity(myseq[i],b0,s0,u0)}
truehist(blame$ptotpalest1[blame$palestpk == 0], nbins=20, xlab = "ptotpalest1", ylab = "pdf", xlim = c(-u0,1+u0))
lines(myseq,mydensity, lwd = 2)
#
# Draw the fitted distribution and histogram for the “unpacked”
for (i in 1:length(mydensity)) {mydensity[i] <- dnsity(myseq[i],b0+b1,s0,u0)}
truehist(blame$ptotpalest1[blame$palestpk == 1], nbins=20, xlab = "ptotpalest1", ylab = "pdf", xlim = c(-u0,1+u0))
lines(myseq,mydensity, lwd = 2)
```

Now let's try the same model with maximum likelihood estimation. We find that the results are quite similar to those obtained with Bayesian estimation.

```{r, eval = FALSE}
# Run the model
m4b <- exftcdfqMLE(ptotpalest1 ~ palestpk | 1 | 1, data = blame)
summary(m4b)
#
# Get the model estimated medians for the “unpacked” and "packed" conditions
predict(m4b, newdata = c(0, 1))
#
# Examine the parameter estimate correlation matrix:
parameter_cor(m4b)
#
```

# References

Halevy, N., Maoz, I., Vani, P. and Reit, E.S., 2022. Where the blame lies: Unpacking groups into their constituent subgroups shifts judgments of blame in intergroup conflict. Psychological Science, 33(1), pp.76-89.

Smithson, M. and Shou, Y., 2024. Flexible CDF-quantile distributions on the closed unit interval, with software and applications. Communications in Statistics-Theory and Methods, 53(11), pp.3876-3898.

Yoon, P., I. Steiner, and G. Reinhardt. 2003. Analysis of factors influencing length of stay in the emergency department. CJEM 5 (3):155–61.
